### 2025-05-19 conclusions.tex
init commit
Modified Ch. Introduction
Almost done report

\chapter{Central Limit Theorem}
\label{clt}

asldfjkldasjflkdfjldsa;f

% \begin{figure*}
% \centering
% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
% \caption{caption}
% \label{figCh1:figure}
% \end{figure*}

\renewcommand\thefigure{\thechapter .\arabic{figure}}

>>>>

%%% -*- Mode: LaTeX; -*-
\chapter{Conclusion}
\label{concl}

We have formalized in this thesis, to a great extent, a general version of the Central Limit Theorem (CLT) in HOL4 under Lyapunov's condition, with the Lindeberg replacement method. The version of the CLT proved herein is far more general than the classical i.i.d. case: it treats summands which are independent but not necessarily identically distributed, with variances and third absolute moments which are not necessarily constant over indices. This is in line with Chung's analytical development \cite{Chung:2001} and is a much stronger version of the theorem than those formalised in theorem provers such as Isabelle/HOL previously \cite{serafin2015formally}.

To our knowledge, this is the first mechanised proof of Lyapunov’s CLT under general independent sequences using Lindeberg replacement method, Taylor expansion and Big-O bounds in HOL4. Compared to the CLT formalized in Isabelle, which assumes the same distribution, our presentation deals with the additional challenge of summands being inhomogeneous. This introduces some technical awkwardness, especially in managing variable-specific error terms in Lindeberg replacement. The result is a version of the CLT valid under Lyapunov's condition—a broader and more applicable class—at the cost of much more intricate formal proofs.

The work involved developing over 6000 lines of HOL4 code, from moment inequalities, summability, Taylor approximation, and asymptotic Big-O estimation theorems and definitions. Most of this infrastructure either had to be developed from scratch or redeveloped to suit the requirements of probability formalisation in HOL4. The development significantly extends the scope of HOL4 for conducting convergence in distribution and limit theorem reasoning.

Early progress on the project was hindered by the learning curve of HOL4 and an initial try at proving the CLT through moment-generating functions. Even though this path is textbook typical, it was unworkable in HOL4 since nothing on MGFs and improper integrals was available in the standard libraries. The transition to the Lindeberg substitute method, while more awkward, was successful and allowed all main analytic components of the CLT to be formally rigorised.

\medskip

\textbf{What has been completed:}
\begin{itemize}
    \item Construction of auxiliary sequences of Gaussian random variables;
    \item Moment and variance control by Lyapunov-type inequalities;
    \item Taylor expansion for the sum and remainder bounds;
\item Big-O asymptotic bounds on the total Taylor approximation error.
\end{itemize}

\textbf{What remains:}
The only remaining work is to complete the proof of the final convergence statement. Specifically, it entails formalising the absolute third moment of the normal distribution as an integral expressed in terms of the gamma function. This requires evaluating \( \mathbb{E}[|X|^3] \) for \( X \sim \mathcal{N}(0, \sigma^2) \), which involves improper integrals over \( \mathbb{R} \), as well as properties of special functions such as \( \Gamma(z) \). Unfortunately, the necessary infrastructure for complex-valued integration, special functions, and improper integrals is not yet available in the standard HOL4 libraries.

In addition to the lack of special function formalisation, a more foundational gap remains between the classical Gauss–Riemann-style integration used in traditional probability theory and the fully formal Lebesgue integration developed in HOL4. Many textbook results rely on informal transitions between these two views of integration, which in a formal setting must be rigorously connected. Bridging this gap—formally relating improper Riemann integrals to their Lebesgue counterparts—is essential for justifying many classical computations, and is currently an area of planned future work.

At the time of writing, the final error bound has been cleanly isolated and stated. All supporting lemmas, inequalities, and approximation steps are in place. The remaining step is therefore not conceptual but purely technical: implementing the final integration using extended analysis tools once they become available in HOL4.

\medskip

\noindent
This thesis gives a good basis for further research work on the formalisation of limit theorems in probability theory. As support for complex integration and special functions in HOL4 evolves, it will be possible to complete the whole proof and perhaps extend to CLTs for dependent variables, martingales, or random vectors. The work presented here not only demonstrates that one can mechanise one of the classic results in probability theory but also develops reusable infrastructure for formalising probabilistic proofs with expectations, variances, Taylor approximations, and convergence in distribution.

### 2025-05-19 futurework.tex
init commit
Modified Ch. Introduction
Almost done report

\chapter{Conclusion}
\label{concl}

asdfsdfasdf
\newpage
asdasasfdfdf

% \begin{figure*}
% \centering
% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
% \caption{caption}
% \label{figCh1:figure}
% \end{figure*}

\renewcommand\thefigure{\thechapter .\arabic{figure}}

>>>>

%%% -*- Mode: LaTeX; -*-
\chapter{Future Work}
\label{future}

The formalisation of Lyapunov’s Central Limit Theorem (CLT) in this thesis has been developed to a nearly complete state. All components except the formal evaluation of $\mathbb{E}[|X|^3]$ for Gaussian $X$ have been mechanised. This includes more than 6,000 lines of definitions, lemmas, and theorems in HOL4. The final convergence theorem has been stated and is supported by all necessary lemmas. The remaining step is to combine all of crucial components and formalise the third absolute moment of the normal distribution:

\[
\mathbb{E}[|X|^3] = \sqrt{8/\pi} \cdot \sigma^3.
\]

This involves computing an improper integral for the gamma function, whose realization is a function of advances in real analysis and complex-valued functions formalism. HOL4 presently lacks sufficient support for improper integration over infinite intervals and integration with complex numbers.

The author's subsequent commitment after this thesis is to complete this final proof step by establishing analytic tools necessary. This entails the closure of the gap between classical improper integrals (say, Riemann/Gauss-type) and formal Lebesgue integral in HOL4. Establishing a connection between these two is essential for justifying many classical results in a formal system.

Beyond completing this proof, several natural directions remain for future work:

\subsection*{1. Generalising the Central Limit Theorem}

The version of the CLT formalised here assumes independent (but not identically distributed) summands. Several natural generalisations can be pursued:
\begin{itemize}
  \item \textbf{CLT for Martingales.} Extend the current proof to martingale difference sequences, which generalise independence to adapted stochastic processes \cite{hall2014martingale}.
  \item \textbf{CLT under Weak Dependence.} Formalise convergence under weak dependence or mixing conditions, as outlined in Billingsley~\cite{billingsley2017probability}.
  \item \textbf{Multivariate CLT.} Extend the proof to sequences of random vectors in \( \mathbb{R}^d \), a standard result in multivariate statistics.
  \item \textbf{Berry–Esseen Bounds.} Provide a quantitative rate of convergence, leveraging the moment bounds and error control machinery developed in this thesis. \cite{berry1941accuracy}
\end{itemize}

\subsection*{2. Extending HOL4’s Analysis and Probability Libraries}

Several features are currently missing or underdeveloped in HOL4 that would enable broader formalisation of classical probability:
\begin{itemize}
  \item Formal definitions and properties of the gamma function, and other special functions such as the error function (\texttt{erf}) and beta function.
  \item Extension of the Lebesgue integral to handle improper and infinite-domain integrals.
  \item Support for real-to-complex and complex-to-complex functions (e.g., \( \mathbb{R} \to \mathbb{C} \)), essential for characteristic function approaches and Fourier analysis.
  \item Higher-level automation for asymptotic approximations, such as reasoning with Big-O notation directly over sequences and sums.
\end{itemize}

This work shows that it is possible not only to formalise advanced results in probability theory, but also to build reusable, modular libraries that can support future theorems in stochastic processes, statistics, and applied machine learning. In conclusion, this thesis provides a good foundation for extensions and establishes the feasibility of mechanizing high-level probability theory in HOL4.

### 2025-05-19 introduction.tex
Almost done report

%%% -*- Mode: LaTeX; -*-
\chapter[Introduction]{Introduction}
\label{chap:1}

The idea of a formalised mechanist is simple but powerful: if we want to trust mathematics in software, we must mechanise not only calculations but also proofs. A mechanist builds a logic engine where every theorem arises from a small set of primitive rules. If those rules are sound, then everything derived from them inherits that soundness.

HOL4 supports this mechanised philosophy~\cite{slind2008brief}. It encodes logic using a typed \(\lambda\)-calculus and enforces soundness by restricting theorem creation to a protected MetaLanguage (ML) type. The only way to construct a theorem is by applying primitive inference rules written in Standard ML. This ensures that the proof kernel checks every line of reasoning with absolute rigour.

In this thesis, we formalise one of the most important results in probability theory: the Central Limit Theorem (CLT). When many independent effects accumulate, their combined influence often resembles a bell curve. We observe this phenomenon across a wide range of domains—from biological traits and human measurements to economic indicators and communication errors. The CLT explains this universality.

We follow the version of the Central Limit Theorem stated in~\cite{chung2000course} and proved using Lindeberg’s method. Let $\{X_j\}_{j=1}^n$ be a sequence of independent random variables such that:

$$
\mathbb{E}[X_j] = 0,\quad \mathbb{E}[X_j^2] = \sigma_j^2,\quad \mathbb{E}[|X_j|^3] = \gamma_j < \infty.
$$

Define the sum $S_n = \sum_{j=1}^n X_j$, the total variance $s_n^2 = \sum_{j=1}^n \sigma_j^2$, and the total third moment $\Gamma_n = \sum_{j=1}^n \gamma_j$. If the Lyapunov condition

$$
\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0
$$

holds, then the normalised sum \(\frac{S_n}{s_n}\) converges in distribution to the standard normal:

$$
\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
$$

This convergence result explains why the normal distribution emerges from a wide variety of processes, even when individual random variables follow skewed or irregular distributions.

We do not assume probability theory or integration as axioms. Instead, we construct them from first principles in HOL4 using its existing real number, measure theory, and Lebesgue integration libraries. We define random variables, convergence in distribution, and weak limits in this framework. We then prove the CLT using the Lindeberg replacement method and Taylor expansion bounds, following the structure outlined in~\cite{Chung:2001}.
Here is the Lyapunov's form of the Central Limit Theorem in HOL4:
\begin{hol}
  \begin{alltt}
    Theorem central_limit_theorem :
    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X N. prob_space p \(\land\)
    ext_normal_rv N p 0 1 \(\land\)
    (\(\forall\)i. real_random_variable (X i) p) \(\land\)
    (\(\forall\)n. indep_vars p X (\(\lambda\)i. Borel) (count n)) \(\land\)
    (\(\forall\)i. expectation p (X i) = 0) \(\land\)
    (\(\forall\)i. expectation p (\(\lambda\)x. (abs (X i x)) pow 3) < +\(\infty\)) \(\land\)
    (\(\forall\)i. variance p (X i) +\(\infty\)) \(\land\)
    (\(\forall\)i. variance p (X i) \(\ne\) 0) \(\land\)
    (\(\forall\)n. (sqrt (second_moments p X n)) \(\ne\) 0) \(\land\)
    ((\(\lambda\)n. (third_moments p X n) / ((sqrt (second_moments p X n)) pow 3)) --> 0)
    sequentially \({\Leftrightarrow}\)
    ((\(\lambda\)n x. (SIGMA (\(\lambda\)i. X i x) (count n)) / (sqrt (second_moments p X n))) --> N)
    (in_distribution p)
  \end{alltt}
\end{hol}

%-----------------------------------------------------

%This formalisation produces a complete, machine-checked proof of Lyapunov’s Central Limit Theorem. It not only demonstrates the power of HOL4 for probabilistic reasoning, but also enriches its measure-theoretic library with tools necessary for formalising deeper results in probability and statistics.
%\newline
%-----------------------------------------------------

Despite not yet completing the final convergence proof, this thesis represents a significant step toward a fully mechanised proof of Lyapunov’s Central Limit Theorem in HOL4. The author developed over 6000 lines of formalisation code, building most of the necessary measure-theoretic and probabilistic infrastructure. Early progress was slowed by the need to learn the HOL4 system and an initial attempt to formalise the CLT using moment-generating functions—a path that proved impractical due to missing theorems in the HOL4 library at the time. The current approach, based on Lindeberg's method and Taylor expansion, has succeeded in formalising all major components of the proof. The only remaining step—formalising the Taylor error bound  and completing the convergence argument—involves calculating the absolute third moment of the normal distribution. This means formalizing integrals using the gamma function and complex numbers, which are currently not yet fully realized in HOL4. At the time of writing the report, the author is still in the midst of this last step. Even if it is still incomplete at the submission deadline, the overall structure and supporting lemmas are all in place, and the last bound is now a question of careful implementation instead of conceptual difficulty.

\section{Motivation}
The Central Limit Theorem (CLT) is more than just a result in probability theory—it is the mathematical explanation behind the bell curve that appears across nature, science, and data. When we take a large number of independent random variables with finite variance, their properly normalised sum tends toward a normal distribution, regardless of the original distributions of the individual variables~\cite{fischer2011history}.This effect manifests in physical measurements, biological readings, economic results, and even in computation roundoff errors. The CLT illustrates the reasons why such behavior is not accidental but is necessary mathematically under gentle hypotheses.

Although succinctly phrased, the CLT is of very deep mathematical organization. Most textbook proofs rely on informal approximations or skip foundational details. In formal verification, this level of vagueness is unacceptable. We need a version of the CLT where every assumption is explicit, every step justified, and the entire argument checked by machine.

Some existing formal proofs of the CLT, such as the one in Isabelle/HOL~\cite{serafin2015formally}, focus on the independent identical distribution (i.i.d.) case and rely on characteristic functions and the Lévy Continuity Theorem. That approach is elegant, but it depends on complex analysis and Fourier transforms, and does not generalise easily to sums of non-identically distributed variables. Moreover, HOL4’s complex number library is still underdeveloped and lacks many of the theorems needed to support such an approach effectively. Attempting to formalise the necessary complex theory would demand substantial additional effort—likely more than what is needed for the Lindeberg-based approach we follow in this work.

In this thesis, we aim for something different. We want to formalise a version of the CLT that:
\begin{itemize}
\item allows non-identically distributed variables;
\item avoids characteristic functions entirely;
\item uses only measure-theoretic and analytical tools available in HOL4.
\end{itemize}

We follow the proof given in~\cite{Chung:2001}, based on Lindeberg’s replacement trick and Taylor expansion bounds. This approach aligns closely with classical analysis and is modular enough to support future generalisations—such as multidimensional CLTs or stable laws.

By formalising this proof in HOL4, we not only demonstrate that the CLT is derivable in a foundational logical system, but also contribute tools and theorems that strengthen HOL4’s probability theory library for future formalisation work.

\section{Thesis Contributions}
The contribution of this work is the formalisation of Lyapunov's form of the Central Limit Theorem in HOL4, using the Lindeberg replacement strategy and Taylor approximation techniques. This required building up the necessary infrastructure for real-valued random variables, expectations, variances, and convergence in distribution from the ground up.

In particular:

\begin{itemize}
\item We constructed the formal setting for expressing the CLT in Lyapunov form, augmenting HOL4 with normalisation tools, moment summation tools, and standardisation tools.
\item We established the background required to define and establish bounds on expectations using Taylor expansion, including a remainder term of order three.
\item We proved key lemmas for the Lindeberg substitution trick, which summarizes how to replace original variables by Gaussian approximations and bound resulting error.
\item We defined the final convergence goal in terms of convergence in distribution with a well-defined formulation and separation of technical specifications.
\end{itemize}

\section{Structure of the Thesis}
The remainder of this thesis is organised as follows. Chapter 2 reviews the Central Limit Theorem, its historical development, and the version we aim to formalise. It also surveys related work, including formal proofs in systems such as Isabelle/HOL. Chapter 3 introduces the HOL4 theorem prover and outlines the relevant libraries and foundational setup. Chapter 4 presents the core of the formalisation, including the supporting lemmas and structure of the proof. Chapter 5 discusses the results and places this work in the context of other efforts in formalised mathematics.Chapter 6 summarises the thesis by stating the contributions and providing guidance for future development of probability theory in HOL4. This research proves that a formalisation of the Central Limit Theorem in HOL4 is possible. It enriches HOL4's probability library, provides reusable tools for future developments, and establishes the system as a platform for fully verified mathematics.

>>>>

\chapter[Introduction]{Introduction}
\label{chap:1}

The idea of a formalised mechanist is simple but powerful: to trust mathematics in software, we must mechanise not only calculations but also proofs. A mechanist builds a logic engine where every theorem arises from a small set of primitive rules. If those rules are sound, everything derived from them inherits that soundness.

HOL4 supports this mechanised philosophy \cite{slind2008brief}. It encodes logic using a typed \(\lambda\)-calculus and restricts theorem creation to a protected MetaLanguage (ML) type, ensuring that every step is checked with mathematical rigour.

In this thesis, we formalise one of the most important results in probability theory: the Central Limit Theorem (CLT). When many independent effects accumulate, their combined influence often resembles a bell curve. The CLT explains this universal phenomenon observed across diverse fields—from biology and economics to physics and computation.

We formalise the Lyapunov version of the CLT from \cite{Chung:2001}, using Lindeberg’s replacement method. Let $\{X_j\}_{j=1}^n$ be a sequence of independent random variables with zero mean, finite variance, and third absolute moment:

\[
\mathbb{E}[X_j] = 0,\quad \mathbb{E}[X_j^2] = \sigma_j^2,\quad \mathbb{E}[|X_j|^3] = \gamma_j < \infty.
\]

Define the sum \( S_n = \sum_{j=1}^n X_j \), total variance \( s_n^2 = \sum \sigma_j^2 \), and total third moment \( \Gamma_n = \sum \gamma_j \). Then, under the Lyapunov condition:

\[
\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0,
\]

we have convergence in distribution:
\[
\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
\]

Here is the Lyapunov's form of the Central Limit Theorem in HOL4:
\begin{hol}
  \begin{alltt}
    Theorem central\_limit\_theorem :
    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X N.
    prob\_space p \(\land\)
    ext\_normal\_rv N p 0 1 \(\land\)
    (\(\forall\)i. real\_random\_variable (X i) p) \(\land\)
    (\(\forall\)n. indep\_vars p X (\(\lambda\)i. Borel) (count n)) \(\land\)
    (\(\forall\)i. expectation p (X i) = 0) \(\land\)
    (\(\forall\)i. expectation p (\(\lambda\)x. (abs (X i x))\(^3\)) < \(+\infty\)) \(\land\)
    (\(\forall\)i. variance p (X i) < PosInf) \(\land\)
    (\(\forall\)i. variance p (X i) \(\ne\) 0) \(\land\)
    (\(\forall\)n. (sqrt (second\_moments p X n)) \(\ne\) 0) \(\land\)
    ((\(\lambda\)n. (third\_moments p X n) / ((sqrt (second\_moments p X n)) pow 3)) --> 0) sequentially
    \(\Rightarrow\)
    ((\(\lambda\)n x. (SIGMA (\(\lambda\)i. X i x) (count n)) / (sqrt (second\_moments p X n))) --> N)
    (in\_distribution p)
  \end{alltt}
\end{hol}

Although the final convergence proof is not yet complete, this thesis represents a significant step toward a full mechanisation. Over 6000 lines of HOL4 code were developed, including formal definitions, theorems, and supporting libraries. An initial attempt using moment-generating functions was abandoned due to insufficient HOL4 support. The current Taylor-based approach has allowed us to formalise all major components. The only missing piece is a bound involving the gamma function, which requires integration techniques not yet supported in HOL4. This is a matter of technical realisation, not conceptual difficulty.
